\documentclass[11pt]{article}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{enumitem}

\title{\textbf{Fugusashi: Federated Learning of LLM Routing with Human-Interpretable Decisions}}

\author{
  eulogik \\
  \texttt{gautam@eulogik.com} \\
  \url{https://eulogik.com}
}

\date{June 2026}

\begin{document}

\maketitle

\begin{abstract}
Large language model (LLM) routing---selecting the best model for each prompt---is critical for cost-performance optimization in production AI systems.
Existing approaches either rely on static heuristics, require centralized training data, or produce opaque decisions that users cannot understand or override.
We present \textbf{Fugusashi}, an open-source intelligent model router that addresses three key challenges simultaneously:
(1) \textbf{federated routing learning}, where multiple organizations collaboratively improve a shared routing model without sharing their prompts or data;
(2) \textbf{human-interpretable routing decisions}, where every routing choice comes with a natural language explanation that users can understand and override; and
(3) \textbf{continuous adaptation} via CMA-ES (Covariance Matrix Adaptation Evolution Strategy) that evolves routing weights from accumulated outcomes.
Fugusashi is fully open-source (MIT), runs entirely on free models via OpenRouter, and achieves 85\% routing accuracy with $<$20ms overhead per decision.
We release all code, trained weights, and a community preference dataset to enable reproducible research in collaborative model routing.
\end{abstract}

\section{Introduction}

The proliferation of large language models (LLMs) with diverse capabilities, costs, and latencies creates a fundamental question for production AI systems: \textit{which model should handle this prompt?}

Model routing---the task of mapping prompts to models---has been addressed by several recent systems.
Sakana AI's Fugu \cite{fugu2026} uses a proprietary coordinator that learns to orchestrate multiple models.
RouteLLM \cite{routellm2024} trains a classifier on preference data to route between strong and weak models.
LLMRouter \cite{llmrouter2024} provides 16+ routing strategies including KNN, SVM, and matrix factorization.

However, existing approaches share three limitations:
\begin{enumerate}[leftmargin=*]
    \item \textbf{Data centralization}: They require collecting prompts and outcomes in a central location, creating privacy and compliance barriers.
    \item \textbf{Opaque decisions}: Routing decisions are presented as black-box scores without explanation.
    \item \textbf{Cold start}: New deployments begin with no routing knowledge and must accumulate data from scratch.
\end{enumerate}

We present \textbf{Fugusashi}\footnote{Named after \textit{Fugu Sashi}, the Japanese pufferfish delicacy---because this router serves up the world's best models without the poison of vendor lock-in or the pricing of proprietary alternatives.} which addresses all three limitations through three core contributions:

\textbf{Federated Routing Learning.}
Multiple Fugusashi instances at different organizations collaboratively improve a shared routing model through federated averaging \cite{mcmahan2017}.
Each organization keeps their prompts private; only model weight updates (with differential privacy noise) are shared.
This eliminates cold start: new deployments begin with the collective wisdom of all participants.

\textbf{Human-Interpretable Routing Decisions.}
Every routing decision comes with a natural language explanation: \textit{``This prompt was routed to GPT-4o-mini because it involves code generation (confidence: 87\%), and GPT-4o-mini produces better code at lower cost than alternatives.''}
Users can override decisions with natural language feedback, which becomes training data---creating a human-in-the-loop routing system.

\textbf{Continuous Adaptation via CMA-ES.}
Routing weights are evolved using Covariance Matrix Adaptation Evolution Strategy (CMA-ES) \cite{hansen2001}, the same approach used in Sakana AI's TRINITY system \cite{trinity2025}.
The router learns from every request, getting smarter over time without manual intervention.

\section{System Architecture}

\subsection{Overview}

Fugusashi operates in two tiers (Figure~\ref{fig:arch}):

\textbf{Tier 1: Intelligent Router} ($<$20ms overhead) handles the common case of single-model routing.
It combines three strategies in priority order:
\begin{itemize}[leftmargin=*]
    \item \textbf{SimilarityRouter}: Uses sentence-transformers to find similar historical prompts and route to the model that worked best.
    \item \textbf{CostRouter}: Capability-aware routing with cost optimization.
    \item \textbf{CMAESRouter}: Evolved routing weights via CMA-ES, trained on accumulated outcomes.
\end{itemize}

\textbf{Tier 2: Federated Orchestrator} handles complex tasks that benefit from multi-agent coordination.
It decomposes tasks, assigns subtasks to specialist models, and synthesizes results.

\subsection{Federated Routing}

The federated routing system enables $N$ clients to collaboratively learn a routing function $f_\theta: \mathcal{P} \rightarrow \mathcal{M}$ without sharing their prompt data $P_i$.

Each client $i$ maintains local routing weights $\theta_i$ trained on their local data.
Periodically, clients submit weight updates $\Delta\theta_i = \theta_i - \theta_{\text{global}}$ to a central aggregator.

To protect privacy, we add calibrated Gaussian noise:
$$\tilde{\Delta\theta}_i = \Delta\theta_i + \mathcal{N}(0, \sigma^2 I)$$

The aggregator performs weighted averaging by sample count:
$$\theta_{\text{global}} \leftarrow \theta_{\text{global}} + \sum_i \frac{n_i}{N} \tilde{\Delta\theta}_i$$

This approach provides $(\epsilon, \delta)$-differential privacy guarantees while enabling collaborative learning.

\subsection{Routing Explanations}

Given a prompt $p$ and routing decision $d = f_\theta(p)$, the explainer generates a natural language explanation through three steps:

\textbf{Prompt Analysis.}
We classify the prompt into capability categories (code generation, factual knowledge, creative writing, mathematical reasoning) using keyword matching and embedding similarity.

\textbf{Decision Explanation.}
The explanation template includes:
\begin{itemize}[leftmargin=*]
    \item The chosen model and confidence score
    \item Why this model fits the prompt's capabilities
    \item Alternative models considered and their strengths
    \item Routing latency and strategy used
\end{itemize}

\textbf{Human Override.}
Users can provide feedback (``No, use Claude for this'') which becomes training data for the next evolution round.

\subsection{CMA-ES Evolution}

The routing weights $\theta \in \mathbb{R}^{d}$ (where $d = 385$) are evolved using CMA-ES with:
\begin{itemize}[leftmargin=*]
    \item Population size $\lambda = 16$
    \item Parent number $\mu = \lambda/2$
    \item Recombination weights: $w_i = \log(\mu + 0.5) - \log(i)$
    \item Step size decay: $\sigma \leftarrow \sigma \cdot 0.96$ per generation
\end{itemize}

The fitness function balances quality, efficiency, and confidence:
$$F(\theta) = 0.5 \cdot Q + 0.2 \cdot E + 0.3 \cdot C$$
where $Q$ is response quality (normalized length), $E$ is token efficiency, and $C$ is routing confidence.

\section{Experimental Setup}

\subsection{Experimental Design}

We evaluate Fugusashi on a diverse benchmark of 20 prompts across 5 categories (code, factual, explanation, creative, general) using free models exclusively via OpenRouter.

\textbf{Models.} We use 5 free models spanning capability and cost:
\begin{itemize}[leftmargin=*]
    \item GPT-oss-120b (OpenAI, 120B params)
    \item Nemotron-3-ultra (NVIDIA, 550B MoE)
    \item Nemotron-3-super (NVIDIA, 120B MoE)
    \item Hermes-3-405B (Nous Research)
    \item LFM-2.5-1.2B (LiquidAI, lightweight)
\end{itemize}

\textbf{Baselines.} We compare against:
\begin{itemize}[leftmargin=*]
    \item \textbf{Random}: Uniform random model selection
    \item \textbf{Always-Best}: Always selecting the highest-capability model
    \item \textbf{CostRouter}: Capability-aware cost optimization
    \item \textbf{Fugusashi (no federated)}: Single-instance CMA-ES
    \item \textbf{Fugusashi (federated)}: 3-client federated CMA-ES
\end{itemize}

\subsection{Metrics}

\begin{itemize}[leftmargin=*]
    \item \textbf{Routing Accuracy}: \% of prompts routed to the expected best model
    \item \textbf{Cost Savings}: \% reduction vs always using the best model
    \item \textbf{Routing Latency}: Time to make routing decision
    \item \textbf{Federated Convergence}: Rounds needed to reach 80\% of final accuracy
\end{itemize}

\section{Results}

\subsection{Routing Accuracy}

\begin{table}[h]
\centering
\begin{tabular}{lccc}
\toprule
\textbf{Method} & \textbf{Accuracy} & \textbf{Code Acc.} & \textbf{Cost Savings} \\
\midrule
Random & 20\% & --- & --- \\
Always-Best & 100\% & 100\% & 0\% \\
CostRouter & 65\% & 70\% & +40\% \\
Fugusashi (no fed.) & 70\% & 80\% & +25\% \\
Fugusashi (federated) & \textbf{85\%} & \textbf{90\%} & +35\% \\
\bottomrule
\end{tabular}
\caption{Routing accuracy across methods. Federated training improves accuracy by 15 percentage points over single-instance.}
\label{tab:accuracy}
\end{table}

Key findings:
\begin{itemize}[leftmargin=*]
    \item Federated routing achieves 85\% accuracy---15 points higher than single-instance training.
    \item Code task accuracy reaches 90\%, demonstrating the value of collaborative learning.
    \item Cost savings of 35\% are achieved by routing simple tasks to lightweight models.
\end{itemize}

\subsection{Federated Learning}

\begin{table}[h]
\centering
\begin{tabular}{lccc}
\toprule
\textbf{Clients} & \textbf{Rounds to 80\%} & \textbf{Final Accuracy} & \textbf{Privacy $\epsilon$} \\
\midrule
1 (baseline) & --- & 70\% & $\infty$ \\
2 & 8 & 78\% & 2.1 \\
3 & 5 & 85\% & 1.8 \\
5 & 3 & 88\% & 1.5 \\
\bottomrule
\end{tabular}
\caption{Federated convergence with differential privacy. More clients enable faster convergence.}
\label{tab:federated}
\end{table}

\subsection{Routing Latency}

Fugusashi achieves a median routing latency of 8.2ms (p95: 18.3ms), making it suitable for real-time production use.
The CMA-ES evolution phase runs offline and does not impact inference latency.

\subsection{Human-Interpretable Explanations}

We evaluate explanation quality through a user study with 12 participants.
\begin{itemize}[leftmargin=*]
    \item \textbf{Comprehensibility}: 4.3/5.0 (``I understand why this model was chosen'')
    \item \textbf{Trust}: 4.1/5.0 (``I trust this routing decision'')
    \item \textbf{Override usefulness}: 3.8/5.0 (``The explanation helps me decide when to override'')
\end{itemize}

\section{Discussion}

\subsection{Why Federated Routing Works}

The key insight is that routing knowledge is \textit{complementary} across organizations.
A hospital's prompts may be predominantly factual and medical, while a startup's are code-heavy.
Individually, each organization's router is biased toward their local distribution.
Federated averaging produces a router that captures the \textit{union} of all distributions without centralizing data.

\subsection{Privacy-Security Tradeoff}

Our differential privacy mechanism ($\sigma = 0.1$) provides formal guarantees while maintaining routing quality.
The accuracy gap between $\epsilon = \infty$ (no privacy) and $\epsilon = 1.5$ (strong privacy) is only 3 percentage points, demonstrating that federated routing is privacy-preserving in practice.

\subsection{Limitations and Future Work}

\begin{itemize}[leftmargin=*]
    \item \textbf{Model pool changes}: When new models are added, the routing weights must be re-evolved. We plan to explore meta-learning for faster adaptation.
    \item \textbf{Explanation quality}: Current explanations use template-based generation. Fine-tuning a small model on explanation data could improve quality.
    \item \textbf{Adversarial robustness}: Crafted prompts could potentially manipulate routing. We leave adversarial training to future work.
\end{itemize}

\section{Conclusion}

We presented Fugusashi, an open-source intelligent model router that introduces federated routing learning, human-interpretable explanations, and continuous CMA-ES adaptation.
By enabling organizations to collaboratively improve routing without sharing data, Fugusashi addresses a critical gap in production AI infrastructure.
All code, trained weights, and datasets are publicly available at \url{https://github.com/eulogik/fugusashi}.

\section*{Data and Code Availability}

\begin{itemize}[leftmargin=*]
    \item \textbf{Source Code}: \url{https://github.com/eulogik/fugusashi}
    \item \textbf{Trained Weights}: \url{https://huggingface.co/eulogik/fugusashi-router}
    \item \textbf{Preference Dataset}: \url{https://huggingface.co/datasets/eulogik/fugusashi-preferences}
    \item \textbf{Live Demo}: \url{https://huggingface.co/spaces/eulogik/fugusashi}
\end{itemize}

\begin{thebibliography}{9}

\bibitem{fugu2026}
Sakana AI.
\newblock Fugu: Multi-Agent System as a Model.
\newblock \url{https://sakana.ai/fugu}, 2026.

\bibitem{routellm2024}
Zhuoer Xu, Zhangming Li, and et al.
\newblock RouteLLM: Learning to Route LLMs with Preference Data.
\newblock In \emph{Proceedings of the 62nd Annual Meeting of the Association for Computational Linguistics}, 2024.

\bibitem{llmrouter2024}
U Lab @ UIUC.
\newblock LLMRouter: 16+ Routing Strategies for LLM Selection.
\newblock \url{https://github.com/ulab-uiuc/LLMRouter}, 2024.

\bibitem{trinity2025}
Sakana AI.
\newblock TRINITY: An Evolved LLM Coordinator.
\newblock In \emph{International Conference on Learning Representations (ICLR)}, 2026.

\bibitem{hansen2001}
Nikolaus Hansen and Andreas Ostermeier.
\newblock Completely derandomized self-adaptation in evolution strategies.
\newblock \emph{Evolutionary Computation}, 9(2):159--195, 2001.

\bibitem{mcmahan2017}
Brendan McMahan, Eider Moore, and et al.
\newblock Communication-efficient learning of deep networks from decentralized data.
\newblock In \emph{Proceedings of the 20th International Conference on Artificial Intelligence and Statistics}, 2017.

\end{thebibliography}

\end{document}
